# Preface {.unnumbered}

```{r setup, include=FALSE}
library(webexercises)
```

<div class="learning-goals">

  - Understand the basics of regression analysis.
  - Learn how to interpret p-values and confidence intervals.
  - Apply regression in R to a real dataset.
</div>

This is a Quarto book.

To learn more about Quarto books visit <https://quarto.org/docs/books>.

```{r, output = "console"}
1 + 1
3+2
```

```{r}
linear_model <- lm(Sepal.Length ~ Sepal.Width, data=iris)
summary(linear_model)
```

## Watch this video

We can add a video fairly easily with the following code (modestbranding reduces logo and rel=0 does not suggest other channels when paused)

(https://quarto.org/docs/authoring/videos.html)

{{< video https://www.youtube.com/embed/iAF7APi_EgU?modestbranding=1&rel=0 >}}

## Callout Boxes

We can make callout boxes for note, warning, important, tip, and caution

(https://quarto.org/docs/authoring/callouts.html)

::: callout-note
Note that there are five types of callouts, including: `note`, `warning`, `important`, `tip`, and `caution`.
:::

::: callout-tip
## Tip with Title

This is an example of a callout with a title.
:::

::: {.callout-caution collapse="true"}
## Expand To Learn About Collapse

This is an example of a 'folded' caution callout that can be expanded by the user. You can use `collapse="true"` to collapse it by default or `collapse="false"` to make a collapsible callout that is expanded by default.
:::

::: {.callout-tip title="Tip with Title"}
This is a callout with a title.
:::

## Sample Questions for Students

::: {.callout-tip title="Try it Out"}
What is the output of `mean(c(1, 2, 3, 4, 5))` in R?

::: {.callout-note collapse="true" title="Solution" appearance="simple"}
The output is `3`. This is because it calculates the arithmetic mean:\
(1 + 2 + 3 + 4 + 5) / 5 = 15 / 5 = 3.
:::
:::

## Creating Problems for Students to Try

::: {.callout-tip title="Try it Out"}
What is the output of `mean(c(1, 2, 3, 4, 5))` in R?

<details>

<summary>Click to see the solution</summary>

The output is `3`. This is because it calculates the arithmetic mean:\
(1 + 2 + 3 + 4 + 5) / 5 = 15 / 5 = 3.

</details>
:::

## Creating Quiz Questions for Students

(https://web.mat.upc.edu/joaquim.puig/posts/webexercises-quiz/)

-   **Numeric questions** How much is 2+3? `r fitb(2+3)`
-   **Multiple Choice** Which is the capital city of Barbados? `r mcq(c(answer="Bridgetown", "Georgetown",             "Kingston", "Bridgerton"))`
-   **True or False** Quarto is very cool `r torf(TRUE)`.
-   **Fill in the Blank** What is the letter after D? `r fitb("E", ignore_case = TRUE)`.
-   Type a vowel: `r fitb(c("A", "E", "I", "O" , "U"), ignore_case = TRUE)`
-   Which function will help me find standard deviation for the code: `r fitb("sd")`(c(1,4,7,8,4))

Testing 123
